```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# paropt

Parameter Optimizing of ODEs

The package *paropt* is build in order to optimize parameters of ode-systems. Thus, the aim is that the output of the integration matches previously measured states. The user has to supply an ode-system in the form of a Rcpp-function. The information about states and parameters are passed via text-files. Additional information such as e.g. the relative tolerance are passed in R.

# Overview

The package *paropt* uses a particle swarm optimizer ('https://github.com/kthohr/optim') in order to find a global best solution. Furthermore, in order to evaluate each particle during optimzation four different solvers can be used all derived from SUNDIALS ('https://computing.llnl.gov/projects/sundials'). For more details see vignette. 

# Example

```{r, echo = F}
path <- system.file("examples", package = "paropt")
setwd(path)
library(paropt)
Rcpp::sourceCpp(paste(path,"/ode.cpp", sep = ""))
df <- read.table(paste(path,"/states_LV.txt", sep = ""), header = TRUE)
time <- df$time
param_start <- "start.txt"
param_lb <- "lb.txt"
param_ub <- "ub.txt"
states <- "states_LV.txt"
state_output <- "final_stateoutput.txt"
par_output <- "optimized_params.txt"
set.seed(1)
optimizer(time, ode, 1e-6, c(1e-8, 1e-8),
         param_start, param_lb, param_ub,
         states, npop = 40, ngen = 200, error = 1,
         state_output, par_output, "bdf")
df_in_silico <- read.table(paste(path, "/final_stateoutput.txt", sep = ""), header = TRUE)
plot(df$time, df$n1, pch = 19, main = "predator", ylab = "predator", xlab = "time")
points(df_in_silico$time, df_in_silico$n1, pch = 19, col = "darkred")
legend(1, 26, legend = c("measured", "in silico"),
col = c("black", "darkred"),lty = 1:2, cex = 0.8) 
plot(df$time, df$n2, pch = 19, main = "prey",ylab = "prey", xlab = "time")
points(df_in_silico$time, df_in_silico$n2, pch = 19, col = "darkred")
legend(1, 26, legend = c("measured", "in silico"),
col = c("black", "darkred"),lty = 1:2, cex = 0.8)
```


# Acknowledgment

I would like to thank all the people who supported me writing this R-Package.
First of all I'm very greatful to my supervisor Arnd Heyer who supported me in writing the package.
Also I would like to thank Satyaprakash Nayak which gave me inspiration to solve coding problems.
Moreover, I would like to thank Tobias Bartsch for elaborately testing of the package.
Furthermore, gratitude is owned to Johannes Krämer for many discussion about simulations. 
Special thanks to my girlfriend Jana for supporting me all the way.

